---
title: "ATSA Final Presentation"
author: "Melanie McCord"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(fpp3)
library(tidyverse)
unemployment_chicago <- readxl::read_excel("data/Chicago-Naperville-ArlingtonHeights_MD_notseasadj.xls", skip = 6) %>%
  drop_na() %>%
  rename(Month = `Month/Year`) %>%
  mutate(`Month` = yearmonth(Month)) %>%
  as_tsibble(index = Month)
```

## Introduction

-  Explore Unemployment in Chicago
-  How has it changed with the pandemic?
-  Can we predict the future?

```{r unemployment_chicago_original}
unemployment_chicago %>%
  autoplot(`Unemployment Rate`)
```

## Unemployment in Chicago



```{r head, echo = TRUE}
colnames(unemployment_chicago)
```

## Characteristics

-  Cyclical pattern (recessions)
-  Seasonality
-  Extreme peak around the start of the pandemic

```{r unemployment_chicago}
unemployment_chicago %>%
  autoplot(`Unemployment Rate`)
```


## Autocorrelation

```{r acf}
unemployment_chicago %>%
  ACF(`Unemployment Rate`) %>%
  autoplot()
```

## Partial Autocorrelation

```{r pacf}
unemployment_chicago %>%
  PACF(`Unemployment Rate`) %>%
  autoplot()
```


## Identifying Appropriate Models

-  Inverse transformation
-  ARIMA and ETS Models
-  Can any of them cope with the changes

## Methodology

-  Training set: data before 2021
-  Automatically selected ARIMA and ETS models
-  Compare with transformed data


## Models

```{r accuracy}
unemployment_chicago <- unemployment_chicago %>%
  mutate(UnemploymentRate = `Unemployment Rate`)
unemployment_chicago_train <- unemployment_chicago %>%
  filter(year(Month) < 2021)
unemployment.fit <- unemployment_chicago_train %>%
  mutate(UnemploymentRate = `Unemployment Rate`) %>%
  model (
    etsinv = ETS(box_cox(UnemploymentRate, lambda = -1)),
    arimainv = ARIMA(box_cox(UnemploymentRate, lambda = -1)),
    arimaauto = ARIMA(UnemploymentRate),
    etsauto = ETS(UnemploymentRate)
  )
unemployment.fit %>%
  pivot_longer(everything(), names_to = "models")
```
# Accuracy
```{r}
unemployment.fit %>%
  forecast(h = "2 years") %>%
  accuracy(unemployment_chicago) %>%
  arrange(RMSE)
```

## CV and AICc Values

```{r}
glance(unemployment.fit)
```

## Plotted Results

```{r fitted}
unemployment.fit %>%
  forecast(h = "2 years") %>%
  autoplot(unemployment_chicago)
```

## Ljung-Box Test on Residuals
```{r}
augment(unemployment.fit) %>%
  features(.innov, ljung_box)
```

## Predicted

```{r pred}
unemployment.fit %>%
  augment() %>%
  autoplot(.fitted) + geom_line(aes(y = UnemploymentRate), color = "black")
```

## ARIMA Residuals
```{r arima}
unemployment.fit %>%
  select(arimaauto) %>%
  gg_tsresiduals()
```


## Conclusions

-  Models were unable to adapt to the rapid changes during the pandemic
-  Unemployment rate in Chicago strongly affected by the pandemic
```{r unemployment_chicago_pred}
unemployment.fit %>%
  select(arimaauto) %>%
  forecast(h = "2 years") %>%
  autoplot(unemployment_chicago)
```

## Future Work

-  Implement knots for the pandemic
-  Add an indicator variable for the pandemic