---
title: "Unemployment in Chicago"
author: "Melanie McCord"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: "show"
bibliography: 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fpp3)
library(tidyverse)
```

#  Introduction

## Motivation

The purpose of this project was to explore unemployment in Chicago and to understand how it changed in relation to the pandemic. The unemployment rate can be affected by the economy, growth in different sectors, as well as other factors. However, the pandemic resulted in major changes to our way of lives, due to closing of in-person events, resulting in major economic changes. In order to understand these trends and how drastic they were, for my project, I chose to explore the unemployment rate in Chicago.

## Dataset

This dataset comes from the non-seasonally adjusted Chicago unemployment metrics from the US bureau of statistics. The period of the data is monthly. I chose to focus on Chicago specifically because the city of Chicago publishes a great deal of metrics and future work aims to extend the analysis of this data into other variables. The time period of the dataset goes from January 1994 to January 2023.

# Data Exploration

```{r}
unemployment_chicago <- readxl::read_excel("data/Chicago-Naperville-ArlingtonHeights_MD_notseasadj.xls", skip = 6) %>%
  drop_na() %>%
  rename(Month = `Month/Year`) %>%
  mutate(`Month` = yearmonth(Month)) %>%
  as_tsibble(index = Month)
head(unemployment_chicago)
```
The data is seasonal, in this case monthly. 

```{r}
unemployment_chicago %>%
  autoplot(`Unemployment Rate`)
```

The data shows clear seasonality, as well as cyclicity in drops and falls of the unemployment rate. There doesn't appear to be any trend in the data, as the sudden growth followed by sharp dips (i.e. economic growth vs. recessions) seems evident in the cyclicity. 

## Exploring the Seasonality

```{r}
unemployment_chicago %>%
  gg_season(`Unemployment Rate`)
```

```{r}
unemployment_chicago %>%
  gg_subseries(`Unemployment Rate`)
```
## STL Decomposition
```{r}
unemployment_chicago %>%
  model (
    STL(`Unemployment Rate`)
  ) %>%
  components() %>%
  autoplot()
```

## Exploring the Autocorrelation

```{r}
unemployment_chicago %>%
  ACF(`Unemployment Rate`) %>%
  autoplot()
```

```{r}
unemployment_chicago %>%
  gg_lag(`Unemployment Rate`, lags = 1:12, geom = "point")
```

```{r}
unemployment_chicago %>%
  PACF(`Unemployment Rate`) %>%
  autoplot()
```

# Identifying Appropriate Models

## Box Cox Transformation
```{r}
unemployment_chicago %>%
  features(`Unemployment Rate`, guerrero)
```

## Fitting an ARIMA and ETS Model

```{r}
unemployment_chicago <- unemployment_chicago %>%
  mutate(UnemploymentRate = `Unemployment Rate`)
unemployment_chicago_train <- unemployment_chicago %>%
  filter(year(Month) < 2021)
unemployment.fit <- unemployment_chicago_train %>%
  mutate(UnemploymentRate = `Unemployment Rate`) %>%
  model (
    etsinv = ETS(box_cox(UnemploymentRate, lambda = -1)),
    arimainv = ARIMA(box_cox(UnemploymentRate, lambda = -1)),
    arimaauto = ARIMA(UnemploymentRate),
    etsauto = ETS(UnemploymentRate)
  )
```

### Forecasting: Visualization and Accuracy

```{r}
unemployment.fit %>%
  forecast(h = "2 years") %>%
  autoplot(unemployment_chicago)
```

```{r}
unemployment.fit %>%
  forecast(h = "2 years") %>%
  accuracy(unemployment_chicago)
```

```{r}
augment(unemployment.fit) %>%
  features(.innov, ljung_box)
```

Although the ETS model performs the best on the future data, the ETS model appears to have very significant autocorrelation in the residuals of the past data, with a p-value of nearly 0. Further, the prediction intervals for the auto ETS data are very wide, stretching to negative numbers, which does not make sense to interpret in context since unemployment rate should be between 0 and 100%.

### Exploring the Residuals
```{r}
unemployment.fit %>%
  select(etsauto) %>%
  gg_tsresiduals()

unemployment.fit %>%
  select(etsauto) %>%
  augment() %>%
  autoplot(UnemploymentRate) + geom_line(aes(y = .fitted), color = "blue")
```

The ACF plot of the residuals of the automatically selected ETS model shows one very significant autocorrelation, and one borderline not statistically significant autocorrelation.

```{r}
unemployment.fit %>%
  select(etsauto) %>%
  gg_tsresiduals()

unemployment.fit %>%
  select(arimaauto) %>%
  augment() %>%
  autoplot(UnemploymentRate) + geom_line(aes(y = .fitted), color = "blue")

unemployment.fit %>%
  select(arimainv) %>%
  augment() %>%
  autoplot(UnemploymentRate) + geom_line(aes(y = .fitted), color = "blue")
```

```{r}
unemployment.fit
```

```{r}
unemployment.fit %>%
  select(etsauto) %>%
  augment() %>%
  autoplot(UnemploymentRate) + geom_line(aes(y = .fitted), color = "blue")
```

Although the ARIMA model fares worse on future data, the residuals do indeed look like white noise, although the normality or skew assumption may be violated due to a few large outliers. It would appear that the ARIMA model may be the better of these models, but overall, none of the models appear to perform that well.

# Conclusions and Future Work

In conclusion, the pandemic drastically changed the unemployment rate in Chicago. Although the ARIMA model was able to capture the past data, it failed to predict future data. The ETS model was closer to the actual data because its default predictions were lower than the ARIMA model, but the ARIMA model overall captured the unemployment data better. The inverse transformation did not improve the results for this particular dataset and in fact, the prediction intervals faired worse because of it.

However, there are some ways to get around that. Future work may include using a piecewise function for the dataset so that the years that the pandemic happened and/or other recessions are considered separately from other years. An issue with this data is that while cyclic recessions are indeed accounted for in other years, the pandemic had a much more drastic effect so that the models struggled to adapt. Additionally, I may consider using an indicator variable for the pandemic and/or other recessions, since in this case it was clear that the cause of the peak in unemployment was related to the pandemic.

Additionally, cross validation on the data exploration may improve the results somewhat since it's easier to predict data less far into the future, but even so, the changes with respect to the pandemic were pretty drastic and the models struggle to adapt.

